Associations of severe COVID-19 with polypharmacy in the REACT-SCOT case-control study (draft only, not to be quoted)

### 2020-06-22 pharmaco.Rmd
updated to use new linkage from 8 June

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 4.4 in those aged under 60 years to 3.5 in those aged 75 years and over.  This association persisted after adjusting for care home residence, any hospital admission in the last five years and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  Partitioning of the "any prescription" variable between BNF chapters, which represent broad indication-based drug classes, showed that prescriptions in drugs in almost all chapters were associated with elevated rate ratios for COVID-19.  We concluded that although most associations of severe COVID-19 with drug prescribing were likely to be attributable to confounding by co-morbidity or frailty, causal effects of drugs on disease susceptibility could not be ruled out.

The objective of this study was to investigate the association of severe COVID-19 with drug prescribing.  Specifically we aimed to test whether this association is accounted for by confounding by co-morbidity and other health-related factors, or by direct effects of prescribed drugs.  A further aim was to determine whether the effects underlying this association are sparse or diffuse.  In the case of causal drug effects, this would correspond to distinguishing between large effects of a few drugs, or an effect of polypharmacy not attributable to any single drug.  In the case of confounding, this would correspond to distinguishing between identifying a few conditions that confer high-risk of COVID-19, or a diffuse effect of multimorbidity.

# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by entry to critical care or death within 28 days of a positive nucleic acid test, regardless of the cause of death given on the death certificate.  The ascertainment of cases and controls is refreshed every few weeks, For this study, the accrued number of cases had increased to `r nrow(cc.severe[cc.severe$CASE==1, ])` For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive [and had not embarked -- to be checked in the CHI database by SW]. Socioeconomic status was encoded as quintile of the postcode-based Scottish Index of Multiple Deprivation (SIMD).

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before the date of testing positive for SARS-CoV-2 so as to include only prescriptions dispensed before the onset of symptoms.  The BNF classification groups drugs by chapter (2 digits), section (2 digits), paragraph (2 digits), subparagraph (1 digit), and chemical substance (2 characters).  BNF subparagraph codes generally represent pharmacological drug classes.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Otherf".  Defined daily doses (DDDs) of each drug were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Average daily doses were calculated for each individual as the sum of (strength $\times$ quantity) over all prescriptions dispensed, divided by (interval length $\times$ DDD), where interval length is the time interval from 1 June 2019 to the cutoff of 15 days before the date that the case (in each stratum of one case and ten matched controls) first tested positive.

# Statistical analysis 
Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`.  As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  ICD-10 and BNF codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565) For this analysis this list was extended to include non-cardiac circulatory disease as analysis of the earlier version of the case-control dataset had shown this to be a risk factor for severe COVID-19.  These co-morbid conditions were assigned based on either a diagnostic code in any hospital discharge record during the last five years, or an dispensed prescription of a drug for which that condition is the only indication.  Diagnoses of diabetes were extracted from linkage to the national diabetes register and supplemented by ICD-10 codes for diabetes and BNF codes for diabetes drugs for recently diagnosed persons not on the last extract of the register .  The ICD-10 codes and drug codes used to define each variable are listed in the R scripts provided as Supplementary Material. <!-- clone the github repo -->

To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we carried out a backward stepwise variable selection procedure. 
Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.  The model with the highest log-likelihood was used to select which drug class should be dropped.  This procedure was repeated with the retained variables until only a single drug class remained.

To study the effect of a specific drug, we constructed a propensity score based on indicator variables for prescriptions of drugs from each of the `r length(subparanames)` BNF subparagraph codes.  As the original linkage of test-positive cases and controls had been done before cases were classified as severe or otherwise, an independent dataset of `r numrows.cchosp` individuals was available to us, comprising test-positive cases who had been hospitalised but had not entered intensive care or died, together with their matched controls. By using this independent dataset to construct the propensity score, we ensured that it would not overfit the case-control dataset.  The propensity score was constructed by stepwise logistic regression with ever-use of the drug under study as dependent variable, and predictor variables as age, sex, and all `r length(subparanames.kept)` subparagraph codes for which at least 5% of case and control records had a dispensed prescription. `r nrow(coeffs.propensity)` drug classes were retained in the propensity score.  As the case-control analysis is conditioned on age and sex, the terms for age and sex were dropped when the score was computed in the case-control dataset.

@@

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.

A backward stepwise variable elimination procedure as described above was carried out to identify which specific drug classes contributed most to the association of COVID-19 with polypharmacy.  To save computation, the stepwise procedure was initiated with the subset of `r length(subparacols.forstepwisedrop)` drug classes in Table \ref{tab:subpara.coeffs} that showed univariate associations.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model.
Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.

@@

* Stratification by propensity score.

## Associations of severe COVID-19 with analgesic use

## Associations of severe COVID-19 with proton pump inhibitor use

Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].  cy.

@@

A propensity score for ever-use of a proton pump inhibitor was constructed as described in the Methods section. The `r nrow(coeffs.propensity)` drug classes retained by stepwise regression, and their coefficients, are shown in Supplementary Table \ref{tab:propensity}.  As expected, the drug classes retained included NSAIDs and antiplatelet drugs, but also other drug classes such as tricyclic antidepressants that are not recognized medical indications for co-prescribing of a proton pump inhibitor. Table \ref{tab:bypropensity} shows that when the case-control dataset was stratified by propensity score, a strong association of COVID-19 with use of proton pump inhibitors remained in those with a propensity score of zero, for whom no prescription had been dispensed for any of the drug classes contributing to the propensity score.

@@

The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements.  Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.  In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small. One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.  We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.  However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.

@@

## Methodological strengths and weaknesses
A strength of this study is that it is based on dispensed prescriptions, rather than issued prescriptions.  Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.  Unless this misclassification of exposed individuals as unexposed differs between those not destined and those destined to develop severe COVID-19, it will lead to underestimation of the rate ratio associated with exposure to this class of drugs and underestimation of the attributable fraction.  Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 classification by trained coders.

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.  The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.  Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors and on the avoidance of inappropriate polypharmacy.

# Declarations
## Information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.  The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)).  Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

@@

```{r numdrugs.agegr, echo=FALSE, warning=FALSE, message=FALSE}

if(FALSE) {
table.numdrugsgr <- rbind(table.numdrugsgr.agegr, table.numdrugsgr.num.icdch)

knitr::kable(table.numdrugsgr,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugsgr",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.agegr)[2:3], 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with number of drug classes (BNF subparagraph codes) dispensed, by age group and by number of ICD chapters represented in hospital diagnoses") %>%
    column_spec(column=1, width="2cm") %>%
    kableExtra::group_rows("Age 0-59", 1, 6) %>%
    kableExtra::group_rows("Age 60-74", 7, 12) %>%
    kableExtra::group_rows("Age 75+", 13, 18) %>%
    kableExtra::group_rows("No admissions", 19, 24, hline_before=TRUE, latex_gap_space="12pt") %>% 
    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 25, 30) %>%
    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 31, 36) %>%
    kable_styling(latex_options="hold_position")
}

```{r numdrugs.num.icdch, echo=FALSE, warning=FALSE, message=FALSE}
knitr::kable(table.numdrugsgr.num.icdch,
             escape=FALSE,
             booktabs=TRUE,
             label="numdrugs.num.icdch",
             row.names=FALSE,
             col.names=c("Number of drug classes", colnames(table.numdrugsgr.num.icdch)[2:3],  
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with polypharmacy in those without listed conditions, by number of ICD-10 chapters represented in hospital diagnoses") %>%
    column_spec(column=1, width="2cm") %>%
    kableExtra::group_rows("No admissions", 1, 6) %>%
    kableExtra::group_rows("1 to 2 ICD-10 chapters in hospital diagnoses", 7, 12) %>%
    kableExtra::group_rows("3 or more ICD-10 chapters in hospital diagnoses", 13, 18) %>%
    kable_styling(latex_options="hold_position")
```

@@

```{r bypropensity, echo=FALSE, warning=FALSE, message=FALSE}

knitr::kable(table.protonpump.bypropensity[, 1:4],
             escape=FALSE,
             booktabs=TRUE,
             label="bypropensity", 
             col.names=c(clean.header(colnames(table.protonpump.bypropensity)[1:2]), 
                         "Rate ratio (95\\% CI)", "$p$-value"),
             caption="Association of severe COVID-19 with ever-use of proton pump inhibitor, by propensity score") %>%
    kable_styling(latex_options="hold_position")

@@

knitr::kable(subpara.coeffs,
             escape=FALSE,
             longtable=TRUE, 
             booktabs=TRUE,
             linesep = linesep, 
             label="subpara.coeffs",
             row.names=FALSE,
col.names=c("BNF subparagraph", colnames(subpara.coeffs)[2:3],  
            "Rate ratio", "$p$-value"),
             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in care homes and without a listed condition, filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
    column_spec(column=1, width="12cm") %>%
    column_spec(column=2, width="1.5cm") %>%
    column_spec(column=3, width="1.5cm") %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()

```{r propensity, echo=FALSE, warning=FALSE, message=FALSE}
chapternums <- gsub("\\..*", "", coeffs.propensity[, 1])
chapternums <- floor(as.integer(chapternums) / 100000)
space.at <- which(diff(chapternums) >0)
linesep.str <- rep('', nrow(coeffs.propensity))
linesep.str[space.at] <- '\\addlinespace'

knitr::kable(coeffs.propensity,
             escape=FALSE,
             label="propensity", 
             booktabs=TRUE,
             digits=c(0, 3), 
             col.names=c("BNF subparagraph", "log odds ratio"),
			 linesep=linesep.str,
             caption="Coefficients of propensity score for prescribing of a proton pump inhibitor, constructed in an independent dataset") %>%
    kable_styling(latex_options="hold_position")

>>>>

# Background
The REACT-SCOT study is a matched case control study of severe COVID-19 in Scotland.  In the first report of this study we reported a strong association of severe COVID-19 with having had at least one prescription dispensed in the past year [@mckeigue_rapid_2020].  The univariate rate ratios associated with at least one prescription varied from 4.4 in those aged under 60 years to 3.5 in those aged 75 years and over.   This association persisted after adjusting for care home residence, hospital admission in the last five years, and diagnoses of conditions designated as high-risk conditions for COVID-19 by the NHS and other public health agencies (hereafter listed conditions).  The objective of this study was to investigate possible explanations for this association: confounding by co-morbidity  factors, or direct effects of prescribed drugs.

# Methods
# Methods
The design of the REACT-SCOT case-control study has been described in detail elsewhere [@mckeigue_rapid_2020].  In brief, all individuals testing positive for nucleic acid for SARS-CoV-2 in Scotland were ascertained through the Electronic Communication of Surveillance in Scotland (ECOSS) database.  Admissions to critical care were obtained from the Scottish Intensive Care Society and Audit Group (SICSAG) database that captures admission to all critical care (intensive care or high dependency) units. Death registrations up to `r format(max(cc.severe$Date.Death, na.rm=TRUE), "%d %B %Y")` were obtained from linkage to the National Register of Scotland.  Severe or fatal COVID-19 was defined by either (1) a positive nucleic acid test followed by entry to critical care or death within 28 days; or (2) a death certificate with COVID-19 as underlying cause.

For each case, the Community Health Index database was used to select ten controls matched for sex, one-year age band and registered with the same primary care practice, who were alive on the same day as the first date that the case tested positive.  For fatal cases who had not tested positive, the incident date was assigned as 14 days before death.  For this study based on ascertainment of test-positive , up to `r format(max(cc.severe$SPECDATE, na.rm=TRUE), "%d %B %Y")` there were `r nrow(cc.severe[cc.severe$CASE==1, ])` cases and `r nrow(cc.severe[cc.severe$CASE==1, ])` controls.

## Morbidity and drug prescribing
For all `r table(cc.severe$CASE)[2]` cases and `r table(cc.severe$CASE)[1]` controls, all ICD-10 diagnostic codes were extracted from the last five years of hospital discharge records in the Scottish Morbidity Record (SMR01), excluding records of discharges less than 25 days before testing positive for SARS-CoV-2, and from the national cancer registry. Diagnoses of diabetes were extracted from linkage to the national diabetes register.  British National Formulary (BNF) drug codes for all prescriptions dispensed since 1 June 2019 were extracted from the Scottish Prescribing Information System [@alvarez-madrazo_data_2016], excluding prescriptions dispensed 15 days or less before testing positive for SARS-CoV-2. The BNF classification groups drugs by seven-digit subparagraph codes that generally correspond to drugs classed by mode of action.  For this analysis prescription codes from chapters 14 and above, comprising dressings, appliances, vaccines, anaesthesia and other preparations were grouped as "Other".  Defined daily doses (DDDs) of each of the five proton pump inhibitors in the BNF were obtained from the [DDD/ATC Index](https://www.whocc.no/atc_ddd_index/) of the WHO Collaborating Centre for Drug Statistics Methodology.  Morphine milligram equivalents (MMEs) for opiates were obtained from the [Faculty of Pain Medicine](https://fpm.ac.uk/opioids-aware-structured-approach-opioid-prescribing/dose-equivalents-and-changing-opioids) of the Royal College of Anaesthetists.  Average daily doses of proton pump inhibitors (as DDDs) and opiates (as MME) were calculated for each individual as the sum of (conversion factor $\times$ strength $\times$ quantity dispensed) divided by (observation period $\times$ DDD).

As described previously, we derived indicator variables for a list of conditions that have been designated as risk conditions for COVID-19 by public health agencies [@nhs_whos_2020].  ICD-10 diagnostic and BNF drug codes used for this are available with the [ENCEPP registration](http://www.encepp.eu/encepp/openAttachment/documents.otherDocument-0/35565).

# Statistical analysis 
Rate ratios for severe COVID-19 (hereafter COVID-19) were estimated from conditional logistic regression models, implemented as Cox regression in the R function `survival::clogit`. To investigate which drug classes contribute to the association of severe COVID-19 with number of drug classes dispensed, we used a backward stepwise variable selection procedure. Starting with a global predictor defined as the sum of indicator variables for all distinct drug classes (BNF subparagraph codes), at each iteration the procedure computed the log-likelihood of models in which each of these drug classes had been dropped before computing the global predictor.  The model with the highest log-likelihood was used to select which drug class should be dropped.  This procedure was repeated with the retained variables until only a single drug class remained.

@@

## Associations of severe COVID-19 with drug classes
We next examined univariate associations of dispensing of each of the `r length(subparacols)` BNF subparagraph codes with COVID-19 in those without listed conditions.  Supplementary Table \ref{tab:subpara.coeffs} shows the univariate rate ratios for the `r length(subparacols.forstepwisedrop)` drug classes that were dispensed to at least 10 cases and were associated with severe disease at $p < 0.001$.  These COVID-19-associated drug classes include widely-prescribed cardiovascular drugs and also multiple drug classes previously identified in Scotland as indicators of overprescribing: laxatives, proton pump inhibitors, hypnotics, anxiolytics, antidepressants, analgesics and nutritional supplements.  
A backward stepwise variable elimination procedure as described above was carried out to identify which of these drug classes contributed most to the association of COVID-19 with polypharmacy.  Figure \ref{fig:stepwise} shows how the log-likelihood changed as these variables were dropped from the model. Dropping the first `r firstrow.neg - 1` drug classes increased the log-likelihood of the model.  Only for the last `r length(rows.print)` drug classes did the log-likelihood decrease when the variable was dropped.  The last drug classes retained by backward selection, and thus the variables with largest independent explanatory power for the effect of polypharmacy, were proton pump, inhibitors, nonopioid /compound analgesics, and opioid analgesics.

@@

## Associations of severe COVID-19 with analgesic use
Table \ref{tab:dosegr.opiate} shows the relationship of COVID-19 to average daily opiate dose (MMEs) over the observation period.  This showed a dose-response relationship.  With those who had not been dispensed opiates as baseline, the univariate rate ratio was XX in those with average dose daily of more than 50 mg morphine equivalent, reduced to YY on adjusting for prespecified covariates.

Table \ref{tab:timewindow.opiate shows that the rate ratio associated with exposure in the most recent 120-day time window was higher than that associated with exposure in a less recent time window (from 1 June 2019 to 121 days before the specimen date)

## Associations of severe COVID-19 with proton pump inhibitor use
Table \ref{tab:dosegr.protonpump} shows that the rate ratio for COVID-19 increases with the dose of proton pump inhibitor, expressed as average DDDs/day over the period for which prescribing was observed. This relationship is strongest in younger age groups, and the effect size is reduced by about one quarter by adjusting for prespecified covariates. 
Of the five drugs in this class in the BNF, only two -- omeprazole and lansoprazole -- were widely prescribed: mean DDDs/day in controls were `r table.ppinames[3, 1]` and `r table.ppinames[2, 1]` respectively.  The effect size for unit increase in dose expressed as average DDDs/day were similar for lansoprazole [DDD 30 mg, rate ratio `r table.ppinames[2, 3]`] and omeprazole rate [DDD 20 mg, rate ratio `r table.ppinames[3, 3]`].  cy.

@@

@@

<!---The propensity score is heavily weighted by other drug classes that are widely over-prescribed but unlikely to have direct effects on susceptibility to COVID-19: antidepressants, analgesics and nutritional supplements. ---> Exposure to proton pump inhibitors is not measured accurately in this study because the time of dispensing does not necessarily correspond to the time of use and because these drugs are also available over the counter.  In this situation, including many covariates in the study that are not directly related to the outcome but are strongly associated with the exposure under study may over-adjust the effect of that exposure towards the null.  By restricting the analysis to those with a propensity score of zero, who had no dispensed prescriptions for any of the `r nrow(coeffs.propensity)` drug classes that contribute to the propensity score, we are able to overcome this but the numbers of exposed cases and controls in this category are relatively small. One of the most powerful pharmacoepidemiologic methods for distinguishing causality from confounding is to compare adverse event rates by time windows of exposure among ever-users of the drug under study.  We did not detect any difference between the rate ratios associated with dispensing of a proton pump inhibitor in more recent and less recent time windows.  However most users of these drugs had prescriptions dispensed in both time windows, and users whose most recent prescription was in the earlier time window may have had a stock of tablets for use as required.

@@

## Methodological strengths and weaknesses
A strength of this study is that it is based on dispensed rather than issued prescriptions.  Our study does not capture all exposure to drugs that are available over the counter, which include proton pump inhibitors.  Although we do not have morbidity data from primary care which would include risk factors such as smoking and obesity, we have diagnoses from hospital discharge records that have been coded to the ICD-10 by trained coders.

## Conclusion
Severe COVID-19 is strongly associated with polypharmacy. This association is not easily explained by co-morbidity, and it is strongest in those without hospital diagnoses of conditions that confer high risk of disease.  The profile of drug prescribing associated with COVID-19 includes many drug classes that are indicators of overprescribing and inappropriate polypharmacy.  The single drug class with the greatest explanatory power for the association of COVID-19 with polypharmacy is proton pump inhibitors.  Although this study did not identify unequivocal evidence of causality for this association, effects of this drug class on susceptibility to enteric infection with SARS-CoV-2 are biologically plausible.  We recommend that public health agencies should reinforce existing guidelines on the overprescribing of proton pump inhibitors, and more generally on the avoidance of unnecessary polypharmacy.

# Declarations
## Information governance
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) that allow Health Protection Scotland staff to link any PHI controlled datasets and amendment 1617-0147 that covers transfer of diabetes data to Health Protection Scotland. Datasets were de-identified before analysis.  The study protocol was registered with the European Network of Centres for Pharmacoepidemiology and Pharmacovigilance (number [EUPAS35558](http://www.encepp.eu/encepp/viewResource.htm?id=35559)). All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on [github](https://github.com/pmckeigue/covid-scotland_public).  Requests for acccess to the data may be submitted to the Public Benefit and Privacy Panel.

@@

@@

@@

knitr::kable(subpara.coeffs,
             escape=FALSE,
             longtable=TRUE, 
             booktabs=TRUE,
             linesep = linesep, 
             label="subpara.coeffs",
             row.names=FALSE,
col.names=c("BNF subparagraph", colnames(subpara.coeffs)[2:3],  
            "Rate ratio", "$p$-value"),
             caption=paste("Univariate rate ratios for severe COVID-19 associated with dispensing of each BNF subparagraph code, restricted to those not resident in care homes and without a listed condition, filtered to show only drug classes with at least 10 exposed cases and $p < 0.001$: ")) %>%
    column_spec(column=1, width="10cm") %>%
    column_spec(column=2, width="1.5cm") %>%
    column_spec(column=3, width="1.5cm") %>%
    kable_styling(latex_options="hold_position") %>%
    landscape()
